A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2024-11-11 sections/03FeatureExtraction.Rmd
config updated

For the central tendency feature of our illustration we chose the robust \textit{median}, which can avoid potential issues with non-normally distributed time series responses or outliers \citep{weisberg1992}. To calculate the \textit{median} ($M$), we let $X_{ij}$ be the ordered list of values from the time series of variable $j$ for participant $i$. The calculation depends on whether the number of measurements in a time series $n$ is odd or even.

```{=tex}
\begin{equation} \label{eq:median}
  M(X_{ij}) = 
    \begin{cases}
      X \left[ \frac{n+1}{2} \right] & \text{if $n$ is odd} \\
      \frac{X \left[ \frac{n}{2} \right] + X \left[ \frac{n}{2} +1 \right]}{2} & \text{if $n$ is even}
    \end{cases}
\end{equation}
```

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018, helmich2021}. Person-level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in between-person differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or do they vary across the measurement period?"

For our illustration data, we chose to capture the time series variability with the \textit{Median Absolute Deviation} ($MAD$), where we calculate the \textit{median} ($M$; calculated as in \equatref{eq:median}) for the absolute deviations of measurement $x$ at time point $t$ for participant $i$ and variable $j$ from the median of that time series $X$. We again chose the robust statistic because the Median-based measure is less affected by non-normal distributions and extreme values or outliers compared to other measures of variability like the standard deviation \citep{weisberg1992}

```{=tex}
\begin{equation} \label{eq:mad}
  MAD(X_{ij}) = M(\left| x_{ijt} - M(X_{ij}) \right|)
\end{equation}
```

\paragraph{Instability.}
Instability captures the average change between two consecutive measurements \citep{ebner-priemer2009}. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the 'jumpy-ness' of the data over time. In other words, variability reflects the range or diversity of values in a time series data, while instability reflects the fluctuation or inconsistency in a time series data over time  \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Within psychological time series, instability measurements have especially been important in the research of borderline personality disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also in understanding early warning signals more generally \citep{wichers2019}. Conceptually, the instability feature, thus, relates to a broad range of research questions, including: "What is the nature of the identification changes in those who start working in a new country?" or "Do strong daily fluctuations in self-esteem reflect the process of identity formation in adolescents?"

@@

Another common measurement of instability is the \textit{Mean of the Squared Successive Differences} ($MSSD$), which is often preferred where differences in magnitude are more important than the frequency of those changes, for example, when big shifts in time series are considered more impactful or when outliers are meaningful and need to be taken into account \citep{chatfield2003}.

\paragraph{Self-similarity.}
Self-similarity in time series data refers to the property of a time series to exhibit similar patterns of behavior over different time scales \citep{dmello2021}. That is, self-similarity describes how much a measurement carries over to future measurements. One important self-similarity in psychological time series is \texit{inertia} --- how much a measurement carries over to its next measurement \citep{kuppens2010, suls1998}. If inertia is high a development tends to stay in a certain state. Because high inertia is resistant to change, in emotion dynamics high inertia of negative affect has been found to be indicative of under-reactive systems and to be characteristic of psychological maladjustment \citep{kuppens2010}. In a similar vein, high inertia in negative affect at baseline was even predictive of the initial onset of depression \citep{kuppens2012}. Conceptually, inertia is more broadly connected to research questions such as: ``Do patients stay in a negative mood for several measurements?'' or ``Do migrants stay with their language practice for several days at a time?''

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead, nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research questions about the type of the development: "Is the development of well-being a nonlinear process?" as well as the shape and structure of the development: "How many spikes in well-being did a migrant experience?"

We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line \citep{wood2017}. The degrees of freedom of a spline model are primarily determined by the number of knots and the order of the spline. For instance, a cubic spline with $k$ knots has $k$+3 degrees of freedom \citep{faraway2016}. However, in a penalized spline framework, which is commonly used for GAMs, the effective degrees of freedom can be less than $k$+3. This is because the model employs a smoothing parameter to control the trade-off between the complexity (flexibility) of the model and its fit to the data, thereby penalizing overly complex models and potentially reducing the effective degrees of freedom \citep{marx1998}. Intuitively then an edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend. The estimated degrees of freedom are commonly based on a concept called 'effective degrees of freedom' and can be represented as the trace $tr$ (i.e., the sum of the diagonal elements) of the smoother matrix $S$, a symmetric matrix that maps from the raw data to the smooth estimates \citep{wood2017}.

@@

Beyond our main features of interest, we also extracted the participant's number of completed ESM measurements to ensure that the clusters are comparable in that regard (i.e., to exclude spurious explanations for the cluster assignments). After the feature extraction, we found that about `r format(round(pMissFull$pmiss %>% filter(set == "All") %>% select(pMissTotal) %>% pull, 2), nsmall=2)`% of the extracted features are missing across the `r ncol(featFull$features %>% select(-ends_with(featureExclusion)))-1` features per participant. This might, for example, happen if participants do not have two subsequent measurements with outgroup interactions, so that an autocorrelation with lag-1 cannot be calculated for the contact-specific variables. The small number of missing values indicates that the feature-based approach indeed largely avoids the structural missingness issue. However, even the few missing values can be an issue for some feature reduction or feature clustering algorithms. We, thus, impute the missing feature values with a single predictive mean matching imputation using the MICE library \citep[][]{buuren2011}. Note again that with this procedure we only need to impute an extremely small number of missing values as most feature calculations can use the available data instead.

>>>>

For the central tendency feature of our illustration, we chose the median (\(M\)), which effectively addresses potential complications arising from non-normally distributed responses or outliers within time series datasets \citep{weisberg1992}. To compute the median, it is imperative to differentiate between two types of time series representations for a given variable \( j \) related to participant \( i \): the chronological series and the ordered series. The chronological time series, denoted by \( X_{ij} \), encapsulates the sequence of observations \( \{x_{ij1}, x_{ij2}, ..., x_{ijn}\} \) for variable \( j \) concerning participant \( i \), organized by their temporal occurrence. Here, \( x_{ijt} \) signifies a specific observation at time \( t \) within this sequence. In contrast, the ordered time series, represented as \( \mathbf{X}_{ij} \), is derived from \( X_{ij} \) by sorting the observations in ascending order of magnitude. This ordered set is expressed as \( \{\mathbf{x}_{ij1}, \mathbf{x}_{ij2}, ..., \mathbf{x}_{ijn}\} \), with each \( \mathbf{x}_{ijk} \) corresponding to the \(k\)-th element in the reordered series \( \mathbf{X}_{ij} \).

The median \( M(\mathbf{X}_{ij}) \) is then the value located precisely at the center of the ordered time series \( \mathbf{X}_{ij} \). Depending on whether the total number of observations (\( n \)) is odd or even, the median is either the middel \(k\)-th element if \( n \) is odd, or the average of the two middle values if \( n \) is even:

```{=tex}
\begin{equation} \label{eq:median}
  M(\mathbf{X}_{ij}) = 
  \begin{cases}
    \mathbf{x}_{ij\left(\frac{n+1}{2}\right)} & \text{if $n$ is odd} \\
    \frac{\mathbf{x}_{ij\left(\frac{n}{2}\right)} + \mathbf{x}_{ij\left(\frac{n}{2} + 1\right)}}{2} & \text{if $n$ is even}
  \end{cases}
\end{equation}
```

This approach ensures that the median is a reliable indicator of central tendency in time series analysis, unaffected by data distribution asymmetries or the presence of outliers.

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. Common measurements of the variability are the variance or standard deviation as well as their robust counter parts. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018, helmich2021}. Person-level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in between-person differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or do they vary across the measurement period?"

For our illustration data, we chose the \textit{Median Absolute Deviation} ($MAD$) to gauge the variability within our time series data. This choice is motivated by the robustness of MAD, particularly its resilience to the effects of non-normal distributions and outliers, which can significantly skew traditional variability measures like the standard deviation \citep{weisberg1992}. For a given variable \( j \) and participant \( i \), the MAD is calculated by first determining the median (\( M \)) of the ordered time series \( \mathbf{X}_{ij} \) as outlined in \equatref{eq:median}. We then compute the absolute deviations of each observation in the time series \( X_{ij} \) from this median value. Specifically, for each time point \( t \), we calculate the absolute difference between \( x_{ijt} \) and the series median \( M(\mathbf{X}_{ij}) \). The MAD is then the median of these absolute deviations:

```{=tex}
\begin{subequations}\label{eq:mad}
    \begin{align} 
      MAD(X_{ij}) &= M(\left| x_{ijt} - M(\mathbf{X}_{ij}) \right|) \label{eq:mad_general} \\
                  &= M(\{ |x_{ij1} - M(\mathbf{X}_{ij})|, |x_{ij2} - M(\mathbf{X}_{ij})|, \ldots, |x_{ijn} - M(\mathbf{X}_{ij})| \}) \label{eq:mad_detailed}
    \end{align}
\end{subequations}
```

The calculation of MAD focuses on the magnitudes of deviations, ensuring it provides a robust measure of dispersion that reflects the inherent variability in the time series data.

\paragraph{Instability.}
Instability captures the average change between two consecutive measurements \citep{ebner-priemer2009, jahng2008}. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the 'jumpy-ness' of the data over time. In other words, variability reflects the range or diversity of values in the un-ordered time series data, while instability reflects the fluctuation or inconsistency in a time series data over time \citep{trull2008, houben2015, koval2013}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep[note that this does not need to be rapidly changing or instable, e.g., when there is linear increase over time; also see][]{jahng2008}. Within psychological time series, instability measurements have especially been important in the research of borderline personality disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also in understanding early warning signals more generally \citep{wichers2019}. Conceptually, the instability feature, thus, relates to a broad range of research questions, including: "What is the nature of the identification changes in those who start working in a new country?" or "Do strong daily fluctuations in self-esteem reflect the process of identity formation in adolescents?"

@@

Another common measurement of instability is the \textit{Mean of the Squared Successive Differences} (\(MSSD\)), which is often preferred where differences in magnitude are more important than the frequency of those changes, for example, when big shifts in time series are considered more impactful or when outliers are meaningful and need to be taken into account \citep{chatfield2003, bos2019}. For psychological ESM data, some research suggests that amplitude and frequency might predict different health outcomes and can be investigated jointly \citep{wang2012a, jahng2008}.

\paragraph{Temporal dependence.}
Temporal dependence in time series data refers to the degree to which a time series is influenced by its past values, exhibiting patterns of behavior that may be regular over different time scales \citep{dmello2021}. In the context of psychological time series, an important aspect of temporal dependence is \textit{inertia} --- how much a measurement carries over to its next measurement \citep{kuppens2010, suls1998}. If inertia is high a development tends to stay in a certain state. Because high inertia is resistant to change, in emotion dynamics high inertia of negative affect has been found to be indicative of under-reactive systems and to be characteristic of psychological maladjustment \citep{kuppens2010}. In a similar vein, high inertia in negative affect at baseline was predictive of the initial onset of depression \citep{kuppens2012}. Conceptually, inertia is more broadly connected to research questions such as: "Do patients stay in a negative mood for several measurements?" or "Do migrants stay with their language practice for several days at a time?"

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead, nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are often best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research questions about the type of the development: "Is the development of well-being a nonlinear process?" as well as the shape and structure of the development: "How many spikes in well-being did a migrant experience?"

We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line \citep{wood2017, bringmann2017}. The degrees of freedom of a spline model are primarily determined by the number of knots and the order of the spline. For instance, a cubic spline with $k$ knots has $k$+3 degrees of freedom \citep{faraway2016, haslbeck2021a, castro-alvarez2024}. However, in a penalized spline framework, which is commonly used for GAMs, the effective degrees of freedom can be less than $k$+3. This is because the model employs a smoothing parameter to control the trade-off between the complexity (flexibility) of the model and its fit to the data, thereby penalizing overly complex models and potentially reducing the effective degrees of freedom \citep{marx1998}. Intuitively then an edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend. The estimated degrees of freedom are commonly based on a concept called 'effective degrees of freedom' and can be represented as the trace $tr$ (i.e., the sum of the diagonal elements) of the smoother matrix $S$, a symmetric matrix that maps from the raw data to the smooth estimates \citep{wood2017}.

@@

\paragraph{Additional considerations.}
Beyond our main features of interest, we also extracted the participant's number of completed ESM measurements to ensure that the clusters are comparable in that regard (i.e., to exclude spurious explanations for the cluster assignments). After the feature extraction, we found that about `r format(round(pMissFull$pmiss %>% filter(set == "All") %>% select(pMissTotal) %>% pull, 2), nsmall=2)`% of the extracted features are missing across the `r ncol(featFull$features %>% select(-ends_with(featureExclusion)))-1` features per participant. This might, for example, happen if participants do not have two subsequent measurements with outgroup interactions, so that an autocorrelation with lag-1 cannot be calculated for the contact-specific variables. The small number of missing values indicates that the feature-based approach indeed largely avoids the structural missingness issue. However, even the few missing values can be an issue for some feature reduction or feature clustering algorithms. We, thus, impute missing feature values via predictive mean matching (PMM) with the MICE package in R, employing a single imputation and specifying a maximum of 50 iterations and a fixed seed for convergence and reproducibility \citep[][]{buuren2011}. We chose PMM for its ability to preserve the original data distribution without assuming normality and robustly handling multiple data types \citep{vanbuuren2006}. Note again that with this procedure we only need to impute an extremely small number of missing values as most feature calculations can use the available data instead.

### 2024-11-11 sections/04FeatureReduction.Rmd
config updated

For our own illustration data, we chose a feature projection method to reduce the dimensionality of our extracted features. We particularly chose the feature projection method for its broad applicability. We, specifically, selected the commonly used \textit{principal component analysis} (PCA). Some of the more tailor-made feature selection algorithms can be more accurate in reducing the feature dimensionality and might retain feature importance information more directly, depending on the specific data structure. However, PCAs have the distinct benefit that they are well-established within the psychometric literature \citep{jolliffe2011} and can broadly be applied to a wide variety of studies in an automatized manner \citep{abdi2010}. As our aim is to present a general illustration that can also be adopted across use cases, we present the workflow using a PCA here but we encourage users to consider more specialized methods as well.

To use the PCA with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration, we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` principal component scores for each participant (i.e., the participants' coordinates in the reduced dimensional space; PC-scores).

We would like to comment on two practical matters when using principal components --- the amount of dimensionality reduction and the interpretation of the principal components. As for the expected dimensionality reduction, given its methodology, PCAs tend to 'work better' at reducing dimensions with (highly) correlated variables \citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous variables and features users will need fewer principal components to explain a large amount of variance, while a more diverse set of variables and features will tend to require more principal components to capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our `r pca_cutoff` principal components are still a relatively high number of variables but this is not surprising as we chose a diverse conceptualization and a diverse set of time series features. As for interpretability, PCA allows users to extract information on the meaning of the principal components. In particular, because the principal components are linear combinations of the original features, users can extract the relative importance of each feature for the extracted principal components (i.e., the eigenvectors). While this can be useful in understanding the variance in the original data or help with manual feature selection, we use the PCA purely to reduce the dimensionality for the clustering step. Instead of relying on the principal components, we use the original features of interest to interpret the later extracted clusters. We particularly advocate for such an approach if all original features are considered meaningful in understanding the time series and users would like to retain the features for interpretation (irrespective of the features' importance).

>>>>

For our own illustration data, we chose a feature projection method to reduce the dimensionality of our extracted features. We particularly chose the feature projection method for its broad applicability. We, specifically, selected the commonly used \textit{principal component analysis} (PCA). Some of the more tailor-made feature selection algorithms can be more accurate in reducing the feature dimensionality and might retain feature importance information more directly, depending on the specific data structure. However, PCAs have the distinct benefit that they are well-established within the psychometric literature \citep{jolliffe2011} and can broadly be applied to a wide variety of studies in an automatized manner \citep{abdi2010}. As our aim is to present a general illustration that can also be adopted across use cases, we present the workflow using a PCA here but we encourage users to consider more specialized methods as well (we provide an example decision guide in \sicontext).

To use the PCA with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally \citep[][]{horne2020}. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration, we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` principal component scores for each participant (i.e., the participants' coordinates in the reduced dimensional space; PC-scores).

We would like to comment on two practical matters when using principal components --- the amount of dimensionality reduction and the interpretation of the principal components. As for the expected dimensionality reduction, given its methodology, PCAs tend to 'work better' at reducing dimensions with (highly) correlated variables \citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous variables and features users will need fewer principal components to explain a large amount of variance, while a more diverse set of variables and features will tend to require more principal components to capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our `r pca_cutoff` principal components are still a relatively high number of variables but this is not surprising as we chose a diverse conceptualization and a diverse set of time series features. As for interpretability, PCA allows users to extract information on the meaning of the principal components. In particular, because the principal components are linear combinations of the original features, users can extract the relative importance of each feature for the extracted principal components (i.e., the eigenvectors). While this can be useful in understanding the variance in the original data or help with manual feature selection, we use the PCA here purely to reduce the dimensionality for the clustering step. Instead of relying on the principal components, we use the original features of interest to interpret the later extracted clusters. We particularly advocate for such an approach if all original features are considered meaningful in understanding the time series and users would like to retain the features for interpretation (irrespective of the features' importance).

### 2024-11-11 sections/06ClusterEvaluationPerformance.Rmd
config updated

kmeansPerf <- list()
for(i in 2:10){
  kmeansPerf[[i-1]] <- fpc::cluster.stats(d = dist(scaled_pc), clustering = kmeansRes[[i-1]]$cluster, G2 = TRUE, G3 = TRUE, aggregateonly = TRUE) %>% unlist
}
kmeansPerfComp <- data.frame(Reduce(rbind, kmeansPerf))

kmeansStatNames <- c(
  "n" = "number of cases", 
  "cluster.number" = "number of points", 
  "min.cluster.size" = "size of smallest cluster", 
  "noisen" = "number of noise points", # Not relevant for kmeans 
  "average.between" = "average distance between clusters", 
  "average.within" = "average distance within clusters", # (reweighted so that every observation, rather than every distance, has the same weight)
  "max.diameter" = "maximum cluster diameter",
  "min.separation" = "minimum cluster separation", 
  "ave.within.cluster.ss" = "within clusters sum of squares", # generalisation
  "avg.silwidth" = "average silhouette width", 
  "g2" = "Goodman Kruskal's Gamma coefficient", # See Milligan and Cooper (1985), Gordon (1999, p. 62). 
  "g3" = "G3 coefficient", # See Gordon (1999, p. 62)
  "pearsongamma" = "Normalized gamma", # correlation between distances and a 0-1-vector where 0 means same cluster, 1 means different clusters. see Halkidi et al. (2001)
  "dunn" = "Dunn index", # minimum separation / maximum diameter, Dunn index, see Halkidi et al. (2002).
  "dunn2" = "Dunn index 2", # minimum average dissimilarity between two cluster / maximum average within cluster dissimilarity, another version of the family of Dunn indexes
  "entropy" = "entropy of distribution of cluster memberships", # see Meila(2007)
  "wb.ratio" = "average within / average between", 
  "ch" = "Calinski and Harabasz index", # (Calinski and Harabasz 1974, optimal in Milligan and Cooper 1985; generalised for dissimilarites in Hennig and Liao 2013)
  "widestgap" = "widest within-cluster gap", 
  "sindex" = "separation index"
)

@@

kmeansOut <- kmeans(scaled_pc, centers = 2, nstart = 100)
```

@@

In our own illustration example, we used the \texttt{cluster.stats()} function from the \texttt{fpc} \textsf{R} package, which calculates a wide variety of internal cluster validity statistics for each of the extracted clustering solutions. With real-world data no single evaluation measure is likely perfect, and different measures may produce different results depending on the characteristics of the data and the research question being addressed \citep{kittler1998}. It is therefore important to consider a variety of evaluation measures and to carefully interpret the results in the context of the specific analysis \citep{vinh2009}. We found that across most indices, the analysis with $k=2$ clusters performed the best. Three commonly reported indices we would like to highlight are the comparison of within clusters sum of squares, the average silhouette score, and the Calinski-Harabasz index. The first statistic we looked at was the total within-cluster sum of square $WCSS$ (see also \equatref{eq:kWCSS}). While the within-cluster variation will naturally decrease with (more) smaller clusters, we observed that the decrease in $WCSS$ was largest until $k=2$ after which the decrease was much smaller. This method is also sometimes referred to as the 'elbow method' \citep{syakur2018}. We then looked at a second, commonly used measure, the average silhouette score. This statistic measures the degree to which each time feature data point is similar to other points within the same cluster, compared to points in other clusters \citep{rousseeuw1987}. In our case, the $k=2$ solution maximized the silhouette coefficient ($s_2=$ `r format(round(kmeansPerfComp$avg.silwidth[kmeansPerfComp$cluster.number==2], 2), nsmall = 2)`). Finally, the Calinski-Harabasz index assesses the compactness and separation of the clusters by assessing the ratio of the sum of between-clusters dispersion and of inter-cluster dispersion for all clusters --- thus, the higher the score the better the performances \citep{calinski1974}. In our case, the $k=2$ solution also showed the highest Calinski-Harabasz index ($CH_2=$ `r format(round(kmeansPerfComp$ch[kmeansPerfComp$cluster.number==2], 2), nsmall = 2)`; a full table of all extracted validity statistics is available in Supplemental Material A)\footnote{It is important to note that another commonly assessed aspect of the evaluation is determining the stability and robustness of the clusters \citep{berkhin2006}. This can be assessed by evaluating the sensitivity of the clusters to different feature sets or clustering algorithms, or by using techniques such as bootstrapping to assess the uncertainty of the clusters \citep{vinh2009}. Especially when comparing different clustering algorithms one common index is the Bayesian information criterion (BIC), where a lower BIC indicates that a model is more representative of the data \citep{vandeschoot2017}.}. In the final $k=2$ solution the k-means analysis also assigned a relatively even number of participants to cluster 1 ($n_{C_1}=$ `r table(kmeansOut$cluster)[1]`) and cluster 2 ($n_{C_1}=$ `r table(kmeansOut$cluster)[2]`).

>>>>

kmeansPerf <- list()
for(i in 2:10){
  kmeansPerf[[i-1]] <- fpc::cluster.stats(d = dist(scaled_pc), clustering = kmeansRes[[i]]$cluster, G2 = TRUE, G3 = TRUE, aggregateonly = TRUE) %>% unlist
}
kmeansPerfComp <- data.frame(Reduce(rbind, kmeansPerf))

kmeansStatNames <- c(
  "n" = "number of cases", 
  "cluster.number" = "number of points", 
  "min.cluster.size" = "size of smallest cluster", 
  "noisen" = "number of noise points", # Not relevant for kmeans 
  "average.between" = "average distance between clusters", 
  "average.within" = "average distance within clusters", # (reweighted so that every observation, rather than every distance, has the same weight)
  "max.diameter" = "maximum cluster diameter",
  "min.separation" = "minimum cluster separation", 
  "ave.within.cluster.ss" = "within clusters sum of squares", # generalisation
  "avg.silwidth" = "average silhouette width", 
  "g2" = "Goodman Kruskal's Gamma coefficient", # See Milligan and Cooper (1985), Gordon (1999, p. 62). 
  "g3" = "G3 coefficient", # See Gordon (1999, p. 62)
  "pearsongamma" = "Normalized gamma", # correlation between distances and a 0-1-vector where 0 means same cluster, 1 means different clusters. see Halkidi et al. (2001)
  "dunn" = "Dunn index", # minimum separation / maximum diameter, Dunn index, see Halkidi et al. (2002).
  "dunn2" = "Dunn index 2", # minimum average dissimilarity between two cluster / maximum average within cluster dissimilarity, another version of the family of Dunn indexes
  "entropy" = "entropy of distribution of cluster memberships", # see Meila (2007)
  "wb.ratio" = "average within / average between", 
  "ch" = "Calinski and Harabasz index", # (Calinski and Harabasz 1974, optimal in Milligan and Cooper 1985; generalised for dissimilarites in Hennig and Liao 2013)
  "widestgap" = "widest within-cluster gap", 
  "sindex" = "separation index"
)

@@

set.seed(12345)
kmeansOut <- kmeans(scaled_pc, centers = 2, nstart = 100)
```

@@

In our own illustration example, we used the \texttt{cluster.stats()} function from the \texttt{fpc} \textsf{R} package, which calculates a wide variety of internal cluster validity statistics for each of the extracted clustering solutions. With real-world data, no single evaluation measure is likely to be perfect. Different measures may yield varying results based on the data characteristics and the research question at hand \citep{kittler1998}. It is therefore important to consider a variety of evaluation measures and to carefully interpret the results in the context of the specific analysis \citep{vinh2009}. We found that across most indices, the analysis with $k=2$ clusters performed the best. Three commonly reported indices we would like to highlight are the comparison of within clusters sum of squares, the average silhouette score, and the Calinski-Harabasz index. The first statistic we looked at was the total within-cluster sum of square $WCSS$ (see also \equatref{eq:kWCSS}). While the within-cluster variation will naturally decrease with (more) smaller clusters, we observed that the decrease in $WCSS$ was largest until $k=2$, after which the decrease was much smaller. This method is also sometimes referred to as the 'elbow method' \citep{syakur2018}. We then looked at a second, commonly used measure, the average silhouette score. This statistic measures the degree to which each time feature data point is similar to other points within the same cluster, compared to points in other clusters \citep{rousseeuw1987}. In our case, the $k=2$ solution maximized the silhouette coefficient ($s_2=$ `r format(round(kmeansPerfComp$avg.silwidth[kmeansPerfComp$cluster.number==2], 2), nsmall = 2)`). Finally, the Calinski-Harabasz index assesses the compactness and separation of the clusters by assessing the ratio of the sum of between-clusters dispersion and of inter-cluster dispersion for all clusters --- thus, the higher the score the better the performances \citep{calinski1974}. In our case, the $k=2$ solution also showed the highest Calinski-Harabasz index ($CH_2=$ `r format(round(kmeansPerfComp$ch[kmeansPerfComp$cluster.number==2], 2), nsmall = 2)`; a full table of all extracted validity statistics is available in Supplemental Material A)\footnote{It is important to note that another commonly assessed aspect of the evaluation is determining the stability and robustness of the clusters \citep{berkhin2006}. This can be assessed by evaluating the sensitivity of the clusters to different feature sets or clustering algorithms, or by using techniques such as bootstrapping to assess the uncertainty of the clusters \citep{vinh2009}. Especially when comparing different clustering algorithms one common index is the Bayesian information criterion (BIC), where a lower BIC indicates that a model is more representative of the data \citep{vandeschoot2017}.}. In the final $k=2$ solution the k-means analysis also assigned a relatively even number of participants to cluster 1 ($n_{C_1}=$ `r table(kmeansOut$cluster)[1]`) and cluster 2 ($n_{C_1}=$ `r table(kmeansOut$cluster)[2]`).

```{r}
#| label: test single cluster solution

library(cluster)

numeric_data <- scaled_pc[, sapply(scaled_pc, is.numeric)]

# also calculate k=1 solution
kmeans_one_cluster <- kmeans(numeric_data, centers = 1, nstart = 25)
kmeans_two_clusters <- kmeans(numeric_data, centers = 2, nstart = 25)

# Compare the total within-cluster sum of squares
total_withinss_one_cluster <- kmeans_one_cluster$tot.withinss
total_withinss_two_clusters <- kmeans_two_clusters$tot.withinss

# Calculate the total sum of squares
total_ss <- sum(sapply(numeric_data, function(x) sum((x - mean(x))^2)))
var_explained_one_cluster <- 1 - (total_withinss_one_cluster / total_ss)
var_explained_two_clusters <- 1 - (total_withinss_two_clusters / total_ss)

# create table output
singleClusterPerf <- data.frame(
  Clusters = c("One Cluster", "Two Clusters"),
  TotalWithinSS = c(total_withinss_one_cluster, total_withinss_two_clusters),
  VarianceExplained = c(var_explained_one_cluster, var_explained_two_clusters)
)
singleClusterPerf$VarianceExplained <-
  ifelse(
    singleClusterPerf$VarianceExplained < 0.001,
    "<.001",
    as.character(round(singleClusterPerf$VarianceExplained, 3))
  )
singleClusterPerfNames <- c("Cluster Count", "Total Within-Cluster SS", "Variance Explained")

# singleClusterPerf %>%
#   kbl(.,
#       escape = TRUE,
#       booktabs = TRUE,
#       align = c("l", "c", "c"),
#       digits = 3,
#       row.names = FALSE,
#       col.names = singleClusterPerfNames,
#       caption = "K-Means Single Cluster Performance Comparison") %>%
#   kable_classic(
#     full_width = FALSE,
#     lightable_options = "hover",
#     html_font = "Cambria"
#   )
```

To ensure that the clustering is necessary in the first place, we also compare the performance to a single cluster solution (i.e., a single centroid). The comparison to this \(k=1\) solution is slightly different because metrics like the between-cluster separation are not available. Nonetheless, comparing the within-cluster sums of squares (SS) and the explained variance, we find that two clusters indeed outperform a single cluster solution. Specifically, the total within-cluster SS decreased from `r format(filter(singleClusterPerf, Clusters == "One Cluster")$TotalWithinSS, digits=2, nsmall=2)` for one cluster to `r format(filter(singleClusterPerf, Clusters == "Two Clusters")$TotalWithinSS, digits=2, nsmall=2)` for two clusters. Additionally, the variance explained increased from `r format(filter(singleClusterPerf, Clusters == "One Cluster")$VarianceExplained, digits=3, nsmall=3)` to `r format(filter(singleClusterPerf, Clusters == "Two Clusters")$VarianceExplained, digits=3, nsmall=3)` when the cluster count was increased to two \citep[e.g., ][; also see \situtorial\ for full results]{beijers2022}.
